D3064R0: Bringing the Laws of Physics, Computer-System Constraints, and Compilers to Bear on the OOTA Problem

### 2024-02-14 ootavacuous.tex
tooling: Add other-problems disclaimer
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

	This paper refactors the commonly used condition for classifying
	OOTA behavior, which permits bringing to bear constraints from
	long-standing laws of physics and from hardware systems.
	In this initial work, we focus on C++ implementations based
	on compilers and computing hardware (including CPUs and GPGPUs).
	We present informal arguments that these constraints and
	limitations prevent OOTA cycles containing only volatile atomics
	from occurring in undefined-behavior-free C++ programs running on
	correctly constructed C++ compiler-based implementations that are
	restricted to single-threaded analysis in our current universe.
	We extend this work to non-volatile atomics by defining
	quasi-volatile behavior that we conjecture that usable real-world
	C++ implementations must adhere to.
\end{abstract}

@@

This paper shows that C++ implementations subject to reasonable
real-world constraints cannot exhibit OOTA cycles.
Section~\ref{sec:Background}
provides background on the problem of OOTA cycles, including a
brief summary of prior work.
Section~\ref{sec:OOTA and Semantic Dependencies}
motivates a shift in definition of OOTA cycle and reviews properties of
semantic dependencies, showing that these properties are a function of
executions rather than simply of the source code.
This leads into
Section~\ref{sec:What is an Execution?},
which describes properties of abstract and hardware executions, along
with the relationship between the two.

This in turn leads to
Section~\ref{sec:C++ Compilers}, which describes properties of the tool
commonly used to create hardware executions with the proper relationship
to a given set of abstract executions, namely  the compiler.
Section~\ref{sec:Hardware Dependencies, Instruction Ordering, and the
Fundamental Property}
then combines hardware dependencies and instruction ordering to arrive
at the fundamental property of semantic dependencies.

This sets the stage for
Section~\ref{sec:Real-World Constraints}
to bring real-world constraints to bear, starting with the laws of
physics, informally demonstrating that these constraints prevent
OOTA cycles containing only volatile and quasi-volatile atomics from
occurring in undefined-behavior-free C++ programs running on correctly
constructed C++ compiler-based implemen- tations that are restricted to
single-threaded analysis in our current universe.

Finally,
Section~\ref{sec:Conclusion}
presents concluding remarks.
% @@@ Future directions at some point.

@@

In broad terms,
OOTA occurs theoretically when a group of threads load from each others' stores
and each thread's store depends on the value returned by that thread's load.
The collection of loads and stores forms an \emph{OOTA cycle}.
In the most extreme cases a nonsensical value can pop up ``out of thin air'',
however, as shown by
Listing~\ref{lst:OOTA Cycle},
OOTA cycles need not involve nonsensical values.

@@

Section~\ref{sec:OOTA: rf vs. rfe}
demonstrates advantages of formulating an OOTA cycle as a cycle in
sdep $\cup$ rfe instead of the traditional sdep $\cup$ rf.
Section~\ref{sec:Properties of Semantic Dependencies}
then discusses properties of semantic dependences, showing that they
are functions of executions rather than strictly of source code.

@@

This section uncovers some semantic-dependency complexities.
Section~\ref{sec:Semantic Dependencies and Source Code}
shows that semantic dependencies are functions of executions,
rather than being strict functions of the source code.
Section~\ref{sec:Semantic Dependencies Can Be Many To One}
shows that semantic dependencies do not necessarily extend
from a single load to a single store, but can instead involve
multiple loads.
Section~\ref{sec:Semantic Dependencies Affected by Cross-Thread Optimizations}
shows that semantic dependencies can be affected by cross-thread
optimizations.
Sections~\ref{sec:Semantic Dependencies Affected by if Statements}
and~\ref{sec:Semantic Dependencies Not Affected by if Statements}
shows that \co{if} statements can have surprising effects on semantic
dependencies, up to and including eliminating them completely.
Finally,
Section~\ref{sec:Semantic Dependencies and Matching Up Stores}
demonstrates some challenges in determining which of a group of
stores is involved in a given semantic dependency.

@@

This section looks more carefully at executions, in some cases revisiting
example code from
Section~\ref{sec:Properties of Semantic Dependencies}.
Section~\ref{sec:Abstract Executions}
looks at executions from the viewpoint of the abstract machine, and
Section~\ref{sec:Hardware Executions}
looks at them from the viewpoint of the computer hardware.
Finally,
Section~\ref{sec:Relation Between Abstract and Hardware Executions}
describes how to relate these two viewpoints.

@@

When a given computer executes the machine code in a file, the outcome
has historically been much better defined than the executions of the
C++ abstract machine.
The hardware's behavior is typically specified with great precision by
the designer or manufacturer, and there are formal, executable memory models
describing exactly what patterns of loads and stores can occur.
Thus, leaving aside questions of asynchronous interrupts and system
calls, the behavior of a CPU executing a particular thread within a
program is entirely determined by the values obtained by the various
load-from-memory instructions.

@@

Taking our cue from the folklore, we propose to recognize formally
that programs with volatile objects can execute in two different kinds
of environment: a benign one in which accesses to these objects work
the same as non-volatile memory accesses, and a non-benign one in which
accesses to volatile objects are subject to outside interference and
act more like I/O.
In particular, when running in a non-benign environment a program's
volatile loads can return unpredictable values.
They don't necessarily read from stores (in contrast to non-volatile loads,
which always must return the value of the store they read from).
This implies that volatile load-acquire's do not necessarily synchronize with
volatile store-releases in the sense of the C++ memory model,\footnote{
	For example, roughly speaking, they might instead synchronize
	with load-acquires and store-releases in device firmware.}
so they do not contribute in the same way to the happens-before relation.
Also, in these environments the rfe relation does not necessarily apply
to volatile loads and hence the accesses in an OOTA cycle must be
non-volatile.

@@

Section~\ref{sec:Users Influence the Behavior of Compilers}
shows that C++ compilers can be influenced by their users as well
as by the standard.
Section~\ref{sec:Global Optimization Can Destroy Dependencies}
presents an example showing that global optimizations can destroy
semantic dependencies, and then
Section~\ref{sec:Inventing Atomic Loads Can Destroy Semantic Dependencies}
analyzes examples showing that inventing atomic loads can destroy
semantic dependencies.
Section~\ref{sec:Volatile and Quasi Volatile Accesses}
then presents required properties of volatile atomic operations, and
also defines quasi-volatile properties for non-volatile atomic operations.

@@

Section~\ref{sec:Dependencies at the Hardware Level}
discusses hardware-level dependencies involving machine instructions.
Section~\ref{sec:Instruction Ordering}
then examines the relationships between various instruction-ordering
mechanisms and semantic dependencies.
This material feeds into
Section~\ref{sec:The Fundamental Property of Semantic Dependencies},
which presents the fundamental property of semantic dependencies
and related complications.

@@

The following sections look at semantic-dependency complications that
can arise in the context of executions.
Section~\ref{sec:Executions Can Be Merged}
shows that executions can be merged,
Section~\ref{sec:Executions Depend on Context}
shows that executions depend on context,
Section~\ref{sec:Executions Can Be Nondeterministic}
looks at nondeterministic executions,
Section~\ref{sec:Executions Can Be Complex}
looks at complex executions,
Section~\ref{sec:Implementers and Users Influence the Definition of Execution}
describes how implementers and users (not just the standard!) can
influence executions,
and finally
Section~\ref{sec:So What Exactly is an Execution, Anyway?}
summarizes this material in preparation for bringing the real world
to bear on these issues.

@@

Because early C~implementers could not come to agreement, standard
does not specify the order of evaluation of function arguments, so the
resulting value of \co{y} might be zero ($\frac{1}{2}$ truncated) or two
($\frac{2}{1}$).
In the former case, there is no semantic dependency from \co{z} to
\co{x}, but in the latter case there is.
Thus, semantic dependencies are not just a function of a particular
execution through the source code, but also of arbitrary choices made
by the C++ implementation.\footnote{
	Again, thanks to Peter Sewell for pointing out this possibility.}

@@

This appendix expands on
Section~\ref{sec:Semantic Dependencies and Tooling}
by looking into the possibility of:

>>>>

	This paper refactors the commonly used condition for classifying
	OOTA behavior, which permits bringing to bear constraints from
	long-standing laws of physics and from hardware systems.
	In this initial work, we focus on C++ implementations based
	on compilers and computing hardware (including CPUs and GPGPUs).
	We present informal arguments that these constraints and
	limitations prevent OOTA cycles containing only volatile atomics
	from occurring in undefined-behavior-free C++ programs running on
	correctly constructed C++ compiler-based implementations that are
	restricted to single-threaded analysis in our current universe.
	We further conjecture that, given current optimizations, these
	constraints also prevent OOTA cycles involving non-volatile
	atomics.
\end{abstract}

@@

@@@ Add brief description and roadmap of paper, including the simplifications
and restrictions.

@@

In broad terms,
OOTA occurs theoretically when a group of threads load from each others' stores
and each thread's store depends on the value returned by that thread's load.
The collection of loads and stores forms an \emph{OOTA cycle}.
In the most extreme cases a nonsensical value can pop up ``out of thin air'',
but this is not necessary.

@@

@@@ Fill in roadmap when sectioning is finalized.

@@

@@@ Roadmap

@@

@@@ Roadmap

@@

When a given computer executes the machine code in a file, the outcome
has historically been much more well defined than the executions of the
C++ abstract machine.
The hardware's behavior is typically specified with great precision by
the designer or manufacturer, and there are formal, executable memory models
describing exactly what patterns of loads and stores can occur.
Thus, leaving aside questions of asynchronous interrupts and system
calls, the behavior of a CPU executing a particular thread within a
program is entirely determined by the values obtained by the various
load-from-memory instructions.

@@

Taking our cue from the folklore, we propose to recognize formally
that programs with volatile objects can execute in two different kinds
of environment: a benign one in which accesses to these objects work
the same as non-volatile memory accesses, and a non-benign one in which
accesses to volatile objects are subject to outside interference and
act more like I/O.
In particular, when running in a non-benign environment a program's
volatile loads can return unpredictable values.
They don't necessarily read from stores (in contrast to non-volatile loads,
which always must return the value of the store they read from).
This implies that volatile load-acquire's do not synchronize with
volatile store-releases in the sense of the C++ memory model,
so they do not contribute in the same way to the happens-before relation.
Also, in these environments the rfe relation does not apply
to volatile loads and hence the accesses in an OOTA cycle must be
non-volatile.

@@

@@@ Fill in roadmap when sectioning is finalized.

@@

@@@ Roadmap

@@

@@

Because early C~implementers could not come to agreement, standard
does not specify the order of evaluation of function arguments, so the
resulting value of \co{y} might be zero ($\frac{1}{2}$ truncated) or two
($\frac{2}{1}$).
In the former case, there is no semantic dependency from \co{z} to
\co{x}, but in the latter case there is.
Thus, semantic dependencies are not just a function of a particular
execution through the source code, but also of arbitrary choices made
by the C++ implementation.\footnote{
	Thanks to Peter Sewell for pointing out this possibility.}

@@

One entirely reasonable reaction is that, given the issues raised in
Sections~\ref{sec:Properties of Semantic Dependencies}
and~\ref{sec:The Fundamental Property of Semantic Dependencies},
load/store reordering is the least of the problems faced by tooling.
Nevertheless, this appendix expands on
Section~\ref{sec:Semantic Dependencies and Tooling}
by looking into the possibility of:
